The interactions of rational, pragmatic agents\\ lead to efficient language structure and use

### 2019-01-30 paper/paper.tex
update with some (not all) of NDGs comments

\begin{abstract}
Languages display a diverse set of distributional regularities such as
the relation between a word's frequency and rank in a corpus, the
distribution of dependency lengths, or the presence of lexical
properties such as ambiguity. We discuss a framework for studying how
these properties emerge from in-the-moment interactions of rational,
pragmatic speakers and listeners. Our work takes Zipfian notions of
lexicon-level efficiency as a starting point, connecting these ideas to
Gricean notions of conversational-level efficiency. To do so, we derive
an objective function for measuring the communicative efficiency of
linguistic systems and then examining the behavior of this objective in
a series of simulations focusing on the communicative function of
ambiguity in language. These simulations suggest that rational pragmatic
agents will produce communicatively efficient systems and that
interactions between such agents provide a framework for examining
efficient properties of language more broadly.

@@

But how does such optimization take place? The example given by Zipf
(1949), describes local, communicative interactions in terms of a
\textit{reference game}. Speakers intend to refer to some object in the
world \(m\). They choose some utterance \(u\) to transmit this intended
meaning, \(u \rightarrow m\). The listener attempts to reconstruct this
intended meaning given the transmitted utterance, \(m \rightarrow u\).
Other functionalist projects have also assumed this basic reference game
setting (Piantadosi et al., 2011; Regier et al., 2015) and this
simplification of the communicative act has proven productive in
theoretical (Ferrer-i-Cancho, 2018), simulation-based (Kirby, Griffiths,
\& Smith, 2014) and empirical explorations (Hawkins, Franke, Smith, \&
Goodman, 2018) of efficient language structure and use.\par

Zipf's conception of speaker and listener effort may be connected to
speakers' pragmatic reasoning in the moment. Under a Gricean treatment
of pragmatics, speakers and listeners are assumed to follow a set of
conversational maxims in which they cooperate to transfer information
(Grice, 1975). These maxims appear to emerge from efficiency concerns,
however (Horn, 1984).\par

In this work, we present a step toward formalizing the framing presented
above -- connecting Zipfian notions of efficient language structure to
Gricean notions of rational conversation (Grice, 1975). To do so
minimally requires three basic ingredients -- (1) a language property
we'd like to explain, (2) a framework for describing in-the-moment
interactions of speaker-listeners, and (3) some measure of linguistic
efficiency. For the current project we focus on the ``communicative
function of ambiguity'' for our property -- translating the
functionalist theory pursued by Piantadosi, et al. (2011) into our
framework. To model local, conversational interactions (2) we adopt the
Rational Speech-act framework (RSA) (Frank \& Goodman, 2012; Goodman \&
Frank, 2016). Finally, (3) we derive an objective function for measuring
language efficiency in the reference game setting.\par

The current work examines the question, ``how does having
\textit{lexical ambiguity} make using language \(\ell\) more or less
useful for communication''? We begin with a high-level introduction to
the modelling framework introducing the basic ingredients we will need
to represent language as repeated reference games. We derive anobjective
function for measuring the efficiency of linguistic systems in this
setting. We move on to two case-studies exploring the questions posed
above -- examining the efficiency of ambiguity from a language design
and language use perspective. The contributions of this work is to show
how the reference game framework, in combination with formal modles of
communication, can be used to study questions at a range of different
timescales, from language use in-the-moment to language change across
historical time.\par

@@

Zipf's example of optimal speaker- and listener-languages took the form
of a reference game. We adopt that formulation here, assuming these
communication games as our basic unit of analysis. In this setting,
speakers and listeners are aware of a set of objects \(M\), which we
will refer to as \emph{meanings} and are knowledgeable about the set of
possible signals \(U\) (\emph{utterances}) that can be used to refer to
a given meaning (see Figure 1). Utterances may have different relative
costs, operationalized via a prior over utterances \(P(U)\). Similarly,
meanings differ in the relative degree to which they need to be talked
about, operationalized as a prior over meanings
\(P(M)\)\footnote{The prior over meanings are analogous to the \textit{need probabilities} assumed in previous work Regier, Kemp \& Kay (2015).}.
We consider a set of contexts \(C\) with an associated prior \(P(C)\).
Contexts describe different orderings of the underlying meaning prior,
formalized as different conditional distributions over meanings e.g.
\(p(M|C=c)\). Finally, we consider a set of communicative events
\(e \in E\) where \(<u, m> = e\) is a tuple of utterance, meaning
pairs.\par

@@

An RSA \emph{speaker agent} defines a conditional distribution over
utterances, mapping from intended meanings \(M\) to utterances \(U\)
using \(\ell\). That is, a speaker defines \(P_{speaker}(u|m;\ell)\). We
will use \(S(u|m;\ell)\) as short-hand throughout. A \emph{listener
agent} defines a conditional distribution over meanings, mapping from
utterances \(U\) to meanings \(M\) using \(\ell\). We will use
\(L(m|u; \ell)\) as shorthand. Note that both speakers and listeners can
induce joint distributions over the set of all events \(E\), although,
importantly, these distributions may differ:
\[P_{speaker}(u, m; \ell) = S(u|m; \ell)p(m)\]
\[P_{listener}(u, m; \ell) = L(m|u; \ell)p(u)\]

@@

Zipf (1949) proposed that the particular distributional properties found
in natural language emerge from competing speaker and listener
pressures. We operationalize this objective in equation (1) -- the
efficiency of a linguistic system \(\ell\) being used by speaker and
listener agents \(S\) and \(L\) is the sum of the expected speaker and
listener effort to communicate over all possible communicative events
\(E\).\par

@@

We assume listener effort is the negative log probability of the
listener's conditional \(-log_2(L(M|U))\). This operationalization of
listener effort is intuitively related to existing work in sentence
processing in which word difficulty during comprehension is proportional
to surprisal (Hale, 2001; R. Levy, 2008).

@@

Finally, we assume that events \(e = <u, m>\) are sampled according the
to following generative model -- some object occurs in the world with
probability \(p(M=m)\), the speaker attempts to refer to that object by
sampling from her conditional distribution \(S(u|m)\) (i.e.
\(e \sim p(m)S(u|m)\)). Using these basic ingredients we arrive at the
following objective:

\begin{equation}
\begin{split}
  = \sum_{u, m}P_{speaker}(u, m; \ell)[-log_2(P_{listener}(u, m; \ell))]
\end{split}
\end{equation}

This equation is the simply the cross-entropy between the speaker and
listener joint distributions.

\begin{equation}
\begin{split}
  = \mathbb{E}_{P_{speaker}}[-log_2(P_{listener})] \\
  = H_{cross}(P_{speaker}, P_{listener}; \ell)
\end{split}
\end{equation}

From an information-theoretic perspective this objective is intuitive:
cross-entropy gives us a measure of dissimilarity between two
distributions -- the average number of bits required to communicate
under one distribution, given that the ``true'' distribution differs. In
our case, this is the difference between the joint distribution assumed
by the speaker \(P_{speaker}\) and listener \(P_{listener}\). In other
words, an ``efficient'' language \(\ell\) minimizes the distance between
what speakers and listener think. For a complete derivation and its
components we refer the reader to Section 2.1 of our supplementary
materials.\par

@@

We provide a toy example of their argument here. Say we have two objects
in the world \(m_1\) and \(m_2\), and two languages \(\ell_1\) and
\(\ell_2\). In language \(\ell_1\), the low-cost utterance \(u_1\) can
be used to refer to both \(m_1\) and \(m_2\) (i.e.
\([[u_1]]_{\ell_1} = \{m_1, m_2 \}\)). With language \(\ell_2\),
however, the low cost utterance \(u_1\) can only refer to \(m_1\) and a
higher-cost utterance \(u_2\) can refer to \(m_2\) (i.e.
\([[u_1]]_{\ell_2} = \{m_1\}\) and \([[u_2]]_{\ell_2} = \{m_2 \}\)).
While it is cheaper for a speaker to use \(\ell_1\) (because she can
always use a low-cost utterance), it is more difficult for a listener
(because \(u_1\) is ambiguous). If context isn't disambiguating then we
might prefer \(\ell_2\) to \(\ell_1\). But, if context is disambiguating
then the speaker can use \(u_1\) to refer to either \(m_1\) or \(m_2\)
and \(\ell_1\) should be preferred.\par

@@

\caption[Optimal languages are more likely to contain ambiguous items as the amount of contextual information increases]{Optimal languages are more likely to contain ambiguous items as the amount of contextual information increases. Vertical axis shows the proportion of optimal languages that contain ambiguity. Horizontal axis shows the number of contexts in each condition (1-4). Red-line represents the optimal language under our Zipfian cross-entropy objective while the blue and red lines show optimal languages under speaker- and listener-only consideration.}\label{fig:plot-optimal-langs}
\end{figure*}
\end{CodeChunk}

@@

We adpot the framework above, and (following Piantadosi et al. 2011),
add \textit{context} (\(C\)). The authors argue that mapping two
meanings \(m_1\) and \(m_2\) to a single utterance \(u_1\) is useful
when they can be disambiguated \emph{in context}. In our case we
consider a context \(c_i \in C\) to specify a permutation of the meaning
probabilities. That is, \(p(m|c_1) \neq p(m|c_2)\), when there are two
contexts \(|C| = \{c_1, c_2\}\).This in turn, leads to an update to our
linguistic efficiency objective as we now would like to consider the
average speaker-listener effort over all contextualized communicative
events (see section 3 SI).\par

@@

Recall that language efficiency is both a function of the particular
semantic mappings induced by that language, the speaker and listener
agents (\(S\) and \(L\)), as well as the utterance (\(P(U)\)), meaning
(\(P(M)\)), and context priors (\(P(C)\)). Rather than assume particular
structure for our utterance, meaning, and context prior distributions,
for each simulation we generate \(P(U) \sim \text{Dir}(1, |U|)\),
\(P(M) \sim \text{Dir}(1, |M|)\), and \(P(C) \sim \text{Dir}(1, |C|)\)
where \(\text{Dir}(1, k)\) specifies the uniform Dirichlet distribution
over a \(k\)-dimensional probability vector.\par

@@

Following the argument given by Piantadosi et al. (2011), we want to
assess the impact of \emph{context} on our objective. To do so we
consider four conditions with \(n=500\) simulations each (that is, 500
unique sets of \(\{P(U), P(M), P(C)\}\). Our first is a
\emph{single-context} condition (\(|C|=1\)) -- there is a only a single
context describing \(P(M)\). Our second condition contains two-contexts
(\(|C| = 2\)) -- we consider efficiency under both \(P(M|C=c_1)\) as
well as \(P(M|C=c_2)\). The third and fourth condition correspond
accordingly with \(|C|=3\) and \(|C| = 4\), respectively.\par

@@

\vspace{1em}

\fbox{\parbox[b][][c]{7.3cm}{\centering All experiments, data, and analysis code are available in the public repository for the project: (link will be available upon acceptance)}}
\vspace{1em} \noindent

@@

\hypertarget{ref-FrankGoodman2012a}{}
Frank, M., \& Goodman, N. (2012). Predicting pragmatic reasoning in
language games. \emph{Science}, \emph{336}, 998.

>>>>

\begin{abstract}
Languages display a diverse set of distributional regularities such as
the relation between a word's frequency and rank in a corpus, the
distribution of dependency lengths, or the presence of lexical
properties such as ambiguity. We discuss a framework for studying how
these properties emerge from in-the-moment interactions of rational,
pragmatic speakers and listeners. Our work takes notions of
lexicon-level efficiency as a starting point, connecting these ideas to
notions of conversational-level efficiency. To do so, we derive an
objective function for measuring the communicative efficiency of
linguistic systems and then examining the behavior of this objective in
a series of simulations focusing on the communicative function of
ambiguity in language. These simulations suggest that rational pragmatic
agents will produce communicatively efficient systems and that
interactions between such agents provide a framework for examining
efficient properties of language more broadly.

@@

But how does such optimization take place? The example given by Zipf
(1949), describes local, communicative interactions in terms of a
\textit{reference game}. Speakers intend to refer to some object in the
world \(m\). They choose some utterance \(u\) to transmit this intended
meaning, \(u \rightarrow m\). The listener attempts to reconstruct this
intended meaning given the transmitted utterance, \(m \rightarrow u\).
Other functionalist projects have assumed this basic reference game
setting (Piantadosi et al., 2011; Regier et al., 2015) and this
simplification of the communicative act has proven productive in
theoretical (Ferrer-i-Cancho, 2018), simulation-based (Kirby, Griffiths,
\& Smith, 2014) and empirical explorations (Hawkins, Franke, Smith, \&
Goodman, 2018) of efficient language structure and use.\par

Zipf's conception of speaker and listener effort may be connected to
speakers' pragmatic reasoning in the moment. Under a Gricean treatment
of pragmatics, speakers and listeners are assumed to follow a set of
conversational maxims in which they cooperate to transfer information
(Grice, 1975). These maxims appear to emerge from efficiency concerns,
however (Horn, 1984). A primary target of the current work is to
formalize this connection, by providing a framework for deriving
system-level efficiencies from local, interaction behavior. Our claim is
that to understand an ``efficient'' property of a system we need to
consider \emph{how} that property is \emph{used} efficiently.
Additionally, any quantitative description of ``efficiency'' should be a
function of both the system and users.\par

In the current work we provide a case-study for the kind of framing
presented above. We choose a property of languages that could, in
principle, vary freely, but shows strong regularities across languages.
The explanadum is why this regularity holds. We examine lexical
ambiguity as our property, extending ideas by Piantadosi et al. (2011).
We define a novel measure of efficiency that depends on the
interactional behavior of interlocutors. For our model of speakers and
listeners we adopt the Rational Speech Act (RSA) framework, which is
nicely attested in experimental data on pragmatic interaction. With
these basic ingredients we have a measure of efficiency given language
and speaker-listener agents. Finally, we show that the property of
interest is prevalent in languages that optimize this measure of
efficiency and also how this property is used efficiently during local
interaction. The primary contribution of this work is to show how the
reference game frameowrk, in combination with formal models of
communication, can be used to connect ideas about system-level
efficiencies to in-the-moment language use.\par

@@

Zipf's example of optimal speaker- and listener-languages took the form
of a reference game. We adopt that formulation here, assuming these
communication games as our basic unit of analysis. In this setting,
speakers and listeners are aware of a set of objects \(M\), which we
will refer to as \emph{meanings} and are knowledgeable about the set of
possible signals \(U\) (\emph{utterances}) that can be used to refer to
a given meaning (see Figure 1). Utterances may have different relative
costs, operationalized via a prior over utterances \(P(U)\). Similarly,
meanings differ in the relative degree to which they need to be talked
about, operationalized as a prior over meanings
\(P(M)\)\footnote{The prior over meanings are analogous to the \textit{need probabilities} assumed in previous work Regier, Kemp \& Kay (2015).}.
We consider a set of contexts \(C\) with an associated prior \(P(C)\).
Contexts encode different meaning priors, formalized as separate
conditional distributions over meanings e.g.
\(p(M|C=c_i) \neq p(M|C=c_j)\). Finally, we consider a set of
communicative events \(e \in E\) where \(<u, m> = e\) is a tuple of
utterance, meaning pairs. Note that in the case of single context
\(|C|=1\) we simply have a prior over meanings \(p(M)\), a
simplification we will use in introducing our objective function.\par

@@

An RSA \emph{speaker agent} defines a conditional distribution over
utterances, mapping from intended meanings \(M\) to utterances \(U\)
using \(\ell\) in a given context \(c\). That is, a speaker defines
\(P_{speaker}(u|m; \ell)\). We will use \(S(u|m; \ell)\) as short-hand
throughout. A \emph{listener agent} defines a conditional distribution
over meanings, mapping from utterances \(U\) to meanings \(M\) using
\(\ell\) in a given context \(c\). We will use \(L(m|u; \ell)\) as
shorthand. Note that both speakers and listeners can induce joint
distributions over the set of all events \(E\), although, importantly,
these distributions may differ:
\[P_{speaker}(u, m; \ell) = S(u|m; \ell)p(m)\]
\[P_{listener}(u, m; \ell) = L(m|u; \ell)p(u)\]

@@

Zipf (1949) proposed that the particular distributional properties found
in natural language emerge from competing speaker and listener
pressures. We operationalize this objective in equation (1) -- the
efficiency of a linguistic system \(\ell\) being used by speaker and
listener agents \(S\) and \(L\) is the sum of the expected speaker and
listener effort to communicate over all possible communicative events
\(E\). Note that to simplify the introduction of this objective we
assume there is just a single context (\(|C|=1\)). In this case
\(p(C=c)=1\) and there is only distribution over meanings
(\(p(m|c) = p(m)\)). We provide a more detailed description of this
derivation in the supplementary materials, both with and without context
\(C\).\par

@@

We assume listener effort is the negative log probability of the
listener's conditional \(-log_2(L(m|u))\). This operationalization of
listener effort is intuitively related to existing work in sentence
processing in which word difficulty during comprehension is proportional
to surprisal (Hale, 2001; R. Levy, 2008).

@@

Importantly, we assume that events \(e = <u, m>\) are sampled according
the to following generative model -- some object occurs in the world
with probability \(p(M=m)\), the speaker attempts to refer to that
object by sampling from her conditional distribution \(S(u|m)\) (i.e.
\(e \sim p(m)S(u|m)\)). From the ingredients highlighted above, it is
possible to derive the following objective between the speaker and
listener joint distributions (see SI 2.2 for a complete derivation).

\begin{equation}
\begin{split}
  = -\mathbb{E}_{P_{speaker}}[log_2(P_{listener})] \\
  = H_{cross}(P_{speaker}, P_{listener}; \ell)
\end{split}
\end{equation}

From an information-theoretic perspective this objective is intuitive:
cross-entropy gives us a measure of dissimilarity between two
distributions -- the average number of bits required to communicate
under one distribution, given that the ``true'' distribution differs. In
our case, this is the difference between the joint distribution assumed
by the speaker \(P_{speaker}\) and listener \(P_{listener}\). In other
words, an ``efficient'' language \(\ell\) minimizes the distance between
what speakers and listeners think.

@@

We provide a toy example of their argument here. Say we have two objects
in the world \(m_1\) and \(m_2\), and two languages \(\ell_1\) and
\(\ell_2\). In language \(\ell_1\), the low-cost utterance \(u_1\) can
be used to refer to both \(m_1\) and \(m_2\) (i.e.
\([[u_1]]_{\ell_1} = \{m_1, m_2 \}\) and
\([[u_2]]_{\ell_1} = \emptyset\)). With language \(\ell_2\), however,
the low cost utterance \(u_1\) can only refer to \(m_1\) and a
higher-cost utterance \(u_2\) can refer to \(m_2\) (i.e.
\([[u_1]]_{\ell_2} = \{m_1\}\) and \([[u_2]]_{\ell_2} = \{m_2 \}\)).
While it is cheaper for a speaker to use \(\ell_1\) (because she can
always use a low-cost utterance), it is more difficult for a listener
(because \(u_1\) is ambiguous). If context isn't disambiguating then we
might prefer \(\ell_2\) to \(\ell_1\). But, if context is disambiguating
then the speaker can use \(u_1\) to refer to either \(m_1\) or \(m_2\)
and \(\ell_1\) should be preferred.\par

@@

\caption[Optimal languages are more likely to contain ambiguous items as the amount of contextual information increases]{Optimal languages are more likely to contain ambiguous items as the amount of contextual information increases. Vertical axis shows the proportion of optimal languages that contain ambiguity. Horizontal axis shows the number of contexts in each condition (1-4). Red-line represents the optimal language under our Zipfian cross-entropy objective while the blue and red lines show optimal languages under speaker- and listener-only objectives.}\label{fig:plot-optimal-langs}
\end{figure*}
\end{CodeChunk}

@@

We adopt the framework above, and (following Piantadosi et al. 2011),
add \textit{context} (\(C\)). The authors argue that mapping two
meanings \(m_1\) and \(m_2\) to a single utterance \(u_1\) is useful
when they can be disambiguated \emph{in context}. In our case we
consider a context \(c_i \in C\) to specify a permutation of the meaning
probabilities. That is, \(p(m|c_1) \neq p(m|c_2)\), when there are two
contexts \(|C| = \{c_1, c_2\}\).This in turn, leads to an update to our
linguistic efficiency objective as we now would like to consider the
average speaker-listener effort over all contextualized communicative
events (see section 3 SI).\par

@@

Recall that language efficiency is both a function of the particular
semantic mappings induced by that language, the speaker and listener
agents (\(S\) and \(L\)), as well as the utterance (\(P(U)\)), meaning
(\(P(M)\)), and context priors (\(P(C)\)). Rather than assume particular
structure for our utterance, meaning, and context prior distributions,
for each simulation we generate \(P(U) \sim \text{Dir}(1, |U|)\),
\(P(M|C=c) \sim \text{Dir}(1, |M|)\) (a separete conditional
distribution over meanings for each context \(c\)), and
\(P(C) \sim \text{Dir}(1, |C|)\) where \(\text{Dir}(1, k)\) specifies
the uniform Dirichlet distribution over a \(k\)-dimensional probability
vector.\par

@@

Following the argument given by Piantadosi et al. (2011), we want to
assess the impact of \emph{context} on our objective. To do so we
consider four conditions with \(n=500\) simulations each (that is, 500
unique sets of \(\{P(U), P(M|C), P(C)\}\). Our first is a
\emph{single-context} condition (\(|C|=1\)) -- there is a only a single
context describing \(P(M)\). Our second condition contains two-contexts
(\(|C| = 2\)) -- we consider efficiency under both \(P(M|C=c_1)\) as
well as \(P(M|C=c_2)\). The third and fourth condition correspond
accordingly with \(|C|=3\) and \(|C| = 4\), respectively.\par

@@

\vspace{1em}
\fbox{\parbox[b][][c]{7.3cm}{\centering All simulation and analysis code are available in the public repository for the project: (link will be available upon acceptance)}}
\vspace{1em} \noindent

@@
